---
title: "Life Expectancy project"
output: html_document
---

By Jinitha. M. J.


USN- 1NT20SDS04

to load the data set
```{r}
df= read.csv("Life expectancy.csv")
```

to read the head of the dataset
```{r}
head(df)
```
To show the number of specified rows i.e for n=8, it will display the first 8 rows

```{r}
head(df, n=8)
```
To display the last Rows of your data
```{r}
tail(df)
```
To view the previously Loaded named object we use ls() command
```{r}
ls()
```
for examining the data structure of an object str() command can be used
```{r}
str(df)
```

to pick out item from third row and third columns:
```{r}
df[3,4]
```

```{r}
class(df$posttest)
```
```{r}
class(dimnames(df))
```
to show a different number of rows 
```{r}
head(df, n=7)
```
to display the bottom 6 lines in the data
```{r}
tail(df)
```
to show a different number of rows
```{r}
tail(df, n=5)
```
To get summary about an Data frame and particular columns  
```{r}
summary(df)
```

to return the structure of the given object with further attributes set
```{r}
head(structure(df))
```




to view all the named objects you have at once
```{r}
ls.str(df)
```

used to narrow down your focus
```{r}
ls.str(pattern = 'df')
```

to view all the columns or rows
```{r}
names(df)

```

to view all the rows
```{r}
row.names(df)
```
to represent names of all the coloumns
```{r}
colnames(df)
```

To see both rows and columns names at the once
```{r}
dimnames(df)
```
to determine the number of items in an object
```{r}
length(df)

```

```{r}
tdf = attach(df)
tdf
```

```{r}
head(stack(df))
```

to get the index
```{r}
head(order(df$Year))
```
to identify the class of a column
```{r}
class(df$Year)
```
Summary of a column
```{r}
summary(df$Year)
```

To find the mean of a column
```{r}
mean(df$Year)
```

To find the median of a column
```{r}
median(df$Year)
```
to find the median absolute deviation
```{r}
mad(df$Year)

```
to find the mode absolute deviation

```{r}
mode(df$Year)
```
to find the variance absolute deviation

```{r}
var(df$Year)
```

to find the sd absolute deviation
```{r}
sd(df$Year)
```
to find the quartile absolute deviation

```{r}
quantile(df$Year)
```

Returns a five number summary 
```{r}
fivenum(df$Year)
```



```{r}
df$Year[3]
```


```{r}
df$Year[1:5]
```
To find the max and min numeric object

```{r}
max(df$Year)
```


```{r}
min(df$Year)
```
sort in ascending order

```{r}
head(sort(df$Year))
```


```{r}
min(df$Life.expectancy)
```


```{r}
head(sort(df$Life.expectancy))
```


 to Compare Two Variances
```{r}
var.test(df$Year, df$Life.expectancy)
```
to get an index

```{r}
head(order(df$Year))
```

to get the rank
```{r}
head(rank(df$Year), n=10)
```
to display the selceted rows and colomns

```{r}
df[3,3]
```


```{r}
df[3, 1:2]
```


```{r}
df[3, 2:3]
```
to covert data frame to matrix

```{r}
head(as.matrix(df))
```
to assemble a matrix/data frames

```{r}
head(rbind(df))

```


```{r}
head(cbind(df))

```

Cumulative Commands
```{r}
head(cumsum(df$Year))

```


```{r}
head(cumsum(df$Life.expectancy))

```


```{r}
head(cummin(df$Year))

```


```{r}
head(cummin(df$Life.expectancy))

```


```{r}
head(cumprod(df$Year))
```

to specify which columns of data you want to use to create your contingency table
```{r}
table(df$Year)

```


```{r}
head(as.table(df$Year))
```

to check if dataframe is extracted
```{r}
is.data.frame(df)
```

```{r}
is.table(df)
```
to form the basis of a logical test

```{r}
if(class(df) == 'data.frame') TRUE else FALSE
```

to creates a simple index
```{r}
seq_along(df)
```
to create a basic stem and leaf plot

```{r}
stem(df$Year)
```


to increase the number of bins

```{r}
stem(df$Year, scale = 2)
```

to create a Histogram 
```{r}
hist(df$Year)
```


```{r}
hist(df$Life.expectancy)

```


```{r}
hist(df$Year, col = 'blue', xlab = 'Year', ylim = c(0, 0.1), freq = FALSE)
```
to find the density

```{r}
density(df$Year)
```

Density Function to Draw a Graph
```{r}
plot(density(df$Year))
```


```{r}
plot(density(df$Life.expectancy))

```


```{r}
hist(df$Year, freq = F)
lines(density(df$Year), lty = 2)
lines(density(df$Year, k = 'rectangular'))
```


```{r}
hist(df$Life.expectancy, freq = F)
lines(density(df$Life.expectancy), lty = 2)
lines(density(df$Life.expectancy, k = 'rectangular'))
```
to generate n random numbers from the normal distribution

```{r}
head(rnorm(df$Year, mean = 0, sd = 1))
  
```

Returns the probability for the quantile q

```{r}
head(pnorm(df$Year, mean = 0, sd = 1))
```
Returns the quantile for a given probability p


```{r}
head(qnorm(df$Year, mean = 0, sd = 1))
```


Gives the density function for values x

```{r}
head(dnorm(df$Year, mean = 0, sd = 1))
```

Quantile-Quantile Plot

```{r}
qqnorm(df$Year)
```

Quantile-Quantile Plot

```{r}

qqnorm(df$Life.expectancy)

```



```{r}
qqnorm(df$Year, main = 'QQ plot of df$Year', xlab = 'Theoretical', 
ylab = 'Quantiles for df$Year')
qqline(df$Year, lwd = 2, lty = 2)
```



```{r}
qqplot(df$Year, df$Life.expectancy)
```



```{r}
qqp = qqplot(df$Year, df$Life.expectancy)
abline(lm(qqp$y ~ qqp$x))
```



```{r}
t.test(df$Year, df$Life.expectancy)
```


```{r}
t.test(df$Year, df$Life.expectancy, var.equal = TRUE)
```


```{r}
t.test(df$Year, mu = 5)

```


```{r}
t.test(df$Life.expectancy, mu = 5)

```

Using Directional Hypotheses
```{r}

t.test(df$Year, mu = 5, alternative = 'greater')
```

Using Directional Hypotheses
```{r}

t.test(df$Life.expectancy, mu = 5, alternative = 'greater')
```

formula Syntax and Subsetting Samples in the t-Test
```{r}

t.test(df$Year ~ df$Entity, data = df, subset = df$Entity %in% c('Australia' , 'Brazil'))
```


```{r}
t.test(df$Year ~ df$Entity, data = df, subset = df$Entity %in% c('Australia' , 'Canada'))
```


```{r}
t.test(df$Year ~ df$Entity, data = df, subset = df$Entity %in% c('Canada' , 'Brazil'))
```

Two-Sample U-Test
```{r}
wilcox.test(df$Year, df$Life.expectancy)
```

basic correlation between x and y

```{r}
cor(df$Year, df$Life.expectancy)
```

Determines covariance between x and y 
```{r}

x = cov(df$Year, df$Life.expectancy)
x
```

to determine the correlation from a matrix of covariance

```{r}
v = as.matrix(x)
cov2cor(v)
```

Significance Testing in Correlation Tests

```{r}
cor.test(df$Year, df$Life.expectancy)
```


```{r}
chisq.test(df$Year
           )
```


```{r}
chisq.test(df$Life.expectancy)
```


```{r}
boxplot(df$Year)
```


```{r}
boxplot(df$Life.expectancy)
```


```{r}
boxplot(df$Year, df$Life.expectancy)
```


```{r}
boxplot(df$Year ~ df$Entity, data = df)
```


```{r}
boxplot(df$Life.expectancy ~ df$Entity, data = df)
```

Horizontal Boxplots
```{r}
boxplot(df$Life.expectancy, horizontal = T)
```

with title
```{r}

boxplot(df$Life.expectancy, horizontal = T)
title(xlab = 'Life expectancies')
```

```{r}
boxplot(df$Year, horizontal = T)
```


```{r}
boxplot(df$Year, horizontal = T)
title(xlab = 'Year')
```



```{r}
library(dplyr)
```


```{r}

temp = df %>% filter(df$Entity == "Australia")
plot(temp$Life.expectancy, temp$Year, pch = 25)
```


```{r}
temp = df %>% filter(df$Entity == "Australia")
plot(temp$Year, temp$Life.expectancy, pch = 25)
```


```{r}
temp = df %>% filter(df$Entity == "Brazil")
plot(temp$Life.expectancy, temp$Year, pch = 25)
```


```{r}
temp = df %>% filter(df$Entity == "Canada")
plot(temp$Year, temp$Life.expectancy, pch = 25)
```


```{r}
with(df, plot(df$Year[df$Entity == 'Brazil'], pch = 0))

```


```{r}
with(df, plot(df$Life.expectancy[df$Entity == 'Canada'], pch = 0))

```



 to get the particular values based on row
```{r}
with(df, (df$Life.expectancy[df$Entity == 'Brazil']))
```


```{r}
with(df, (df$Life.expectancy[df$Entity == 'Canada']))
```


```{r}
with(df, (df$Year[df$Entity == 'Brazil']))
```
to creates a matrix of scatter plots.


```{r}
pairs(~ Life.expectancy + Year, data = temp)
```


```{r}
pairs(~Year  + Life.expectancy, data = temp)

```

Line plot

```{r}

plot(temp$Year, type = 'l' )
```


```{r}
plot(temp$Life.expectancy, type = 'l' )

```


```{r}
plot(temp$Year, type = 'b')
```


```{r}
plot(temp$Life.expectancy, type = 'b')
```
Pie chart

```{r}

pie(temp$Life.expectancy, labels = 'Life expextancy')
```


```{r}
plot(temp$Year, type = 'b')
```


```{r}
pie(temp$Year, labels = 'year')
```
Cleveland Dot Charts

```{r}

dotchart(temp$Year)
```


```{r}
dotchart(temp$Life.expectancy)
```

Bar plot
```{r}

barplot(temp$Year)
title(xlab = 'Australia', ylab = 'Year')
box()
```


```{r}
barplot(temp$Life.expectancy)
title(xlab = 'Brazil
      ', ylab = 'Life expectancy')
box()
```


```{r}
barplot(temp$Year)
title(xlab = 'Brazil', ylab = 'Year')
box()
```

to get the column’s diff level(type) counts
```{r}

table(df$Entity)
```


```{r}
barplot(temp$Life.expectancy, horiz = T)
title(xlab = 'Australia', ylab = 'Life expectancy')
box()
```


```{r}
barplot(temp$Life.expectancy, horiz = T)
title(xlab = 'Canada', ylab = 'Life expectancy')
box()
```


```{r}
barplot(temp$Life.expectancy, horiz = T)
title(xlab = 'Brazil', ylab = 'Life expectancy')
box()
```

the analysis of variance 
```{r}

temp.aov = na.omit(aov(temp$Life.expectancy ~ temp$Year, data = temp))
temp.aov
```


```{r}
temp.aov = na.omit(aov(temp$Year ~ temp$Life.expectancy, data = temp))
temp.aov
```

To see the classic ANOVA table
```{r}

summary(temp.aov)
```


```{r}

factor(df$Entity)
```

creating new testCol
```{r}
testCol = c(rep(df$Year - df$Life.expectancy, length(df)))
head(testCol)


```


```{r}
testRow = c('Australia', 1803, 34.05)
testRow
```


```{r}
testCol = c(rep(df$Life.expectancy - df$Year, length(df)))
head(testCol)
```


```{r}
testRow = c('Spain', 1919, 41.12)
testRow
```

Adding Rows or Columns
```{r}

test = data.frame(df, testCol)
head(test)
```


```{r}
test = df
test['testRow',] = testRow
tail(test)
```


```{r}
test = df
test["Test",] = NA
tail(test)
```

Column Indexes
```{r}

col(temp, as.factor = F)
```

Extracting two(2) columns from df
```{r}
x = df[ , 1:2]
head(x)
```


```{r}
x = df[ , 2:3]
head(x)
```

Simple Column and Row Summaries
```{r}
colMeans(x)
```


```{r}
rowMeans(x)

```


```{r}
colSums(x)
```


```{r}
rowSums(x)
```
to apply a function over all the rows or columns of a data frame

```{r}
apply(df[ , 1:3], 2, mean)
```


```{r}
apply(df[ , 1:2], 1, mean)
```


```{r}
apply(df[ , 1:3], 2, median)
```


```{r}
apply(df[ , 2:3], 2, median)
```


```{r}
apply(df[ , 1:3], 2, var)
```

```{r}
apply(df[ , 1:3], 2, var)
```

Using tapply() to Summarize Using a Grouping Variable
```{r}

tapply(df$Year, df$Entity, FUN = mean)
```


```{r}
tapply(df$Life.expectancy, df$Entity, FUN = mean)
```


```{r}
aggregate(df[ , 1:3], by = list(df$Entity), FUN = mean)
```

uding cbind()
```{r}

aggregate(cbind(df$Life.expectancy, df$Year), data = df, by = list(df$Entity), FUN = mean)
```


```{r}
aggregate(df$Life.expectancy ~ df$Entity, data = df, FUN = mean)
```


```{r}
aggregate(df$Year ~ df$Entity, data = df, FUN = mean)
```
The na.omit() command strips out unwanted NA items from vectors and data frames.
```{r}

head(na.omit(df))
```


```{r}
objects(df)
```


```{r}
ls(df)
```



Simple linear regression
```{r}
cor.test(~ df$Year + df$Life.expectancy, data = df)
```


```{r}
dflm = lm(Year ~ Entity, data = df)
dflm
```

extract the coefficients

```{r}
coef(dflm)

```

obtain confidence intervals

```{r}

confint(dflm)
```

Fitted Values
```{r}

head(fitted(dflm))



```

Residuals
```{r}
head(residuals(dflm))
```

Formula
```{r}
formula(dflm)

```


```{r}
dflm$call
```


```{r}
formula(df)
```
Command for Error Bars
```{r}

df.m = apply(df[,1:2], 2, mean)
df.m

```


```{r}
df.sd = apply(df[ , 1:3], 2, sd)
df.sd
```


```{r}
df.s = apply(df[,3:2], 2, sum)
df.s
```


```{r}
df.1 = df.s/df.m
df.1

```


```{r}
df.se = df.sd / sqrt(df.1)
df.se

```


```{r}
df.m + df.se
```


```{r}
max(df.m + df.se)
```


```{r}
df.max = round(max(df.m + df.se) + 0.5, 0)
df.max
```




To Create Mathematical Expressions
```{r}
plot(1:10, 1:10, type = 'n')
opt = par(cex = 1.5)
text(1, 1, expression(hat(x)))
text(2, 2, expression(alpha==x))
text(3, 3, expression(beta==y))
text(4, 4, expression(frac(x, y)))
text(5, 5, expression(sum(x)))
text(6, 6, expression(sum(x^2)))
text(7, 7, expression(bar(x) == sum(frac(x[i], n), i==1, n)))
text(8, 8, expression(sqrt(x)))
text(9, 9, expression(sqrt(x, 3)))
```









